\subsection*{Defining data elements}

\input{schema-figure.tex} 

Consider a particular real-world neuroscience study comprising a series of experiments that involve simultaneous \emph{in vivo} recordings of several types of physiological signals (whole-cell membrane potential or V\textsubscript{m}, local-field potentials or LFP, and calcium fluorescence signals), stimuli (visual display and optogenetic stimulation of targeted neuronal populations), and behaviors (locomotion, whisking, eye movements) \cite{reimer_pupil_2014}. 
Figure \ref{schema}\,A illustrates such an experiment.

To work with data from such studies, DataJoint users create a \emph{schema} (or several schemas) comprising a collection of \emph{base relations} to represent the various elements of the experiments (Fig.\ \ref{schema}\,B).
Relations are DataJoint's basic data representation and can be thought of as simple tables with a \emph{heading} and a \emph{body}.
The heading specifies attribute names and datatypes. 
The body comprises a set of \emph{tuples} of attribute values. 
Base relations are stored in the database whereas \emph{derived relations} may be constructed from base relations for data queries.
For detailed definitions of the Relational Data Model, see Table \ref{glossary}.

\input{glossary.tex}

A base relation is created in the form of a class in MATLAB or Python (Fig.\ \ref{schema}\,C) that defines the relation's heading.
The heading definition comprises a description of the relation, dependencies on other base relations, and a set of \emph{attributes}. 
Each attribute has a name, an optional default value, a datatype, and an optional description.
Attributes that comprise the \emph{primary key} of the relation are separated from the remaining attributes by the divider {\tt---}.

DataJoint provides all the functionality for accessing and manipulating data through the base relation classes. 

